---
title: "Brockett soil C regression kriging"
author: "Mike Whitfield"
date: "21/07/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set up required packages and functions

The following packages and functions are required to run this analysis
```{r load packages functions, echo=FALSE}
library(rgdal)
library(geoR)
library(gstat)
library(dplyr)
library(tidyr)
library(here)
library(automap)
library(readr)
library(knitr)

## Function from Stackoverflow to remove rows containing missing data from spatial objects
# FUNCTION TO REMOVE NA's IN sp DataFrame OBJECT
#   x           sp spatial DataFrame object
#   margin      Remove rows (1) or columns (2) 
sp.na.omit <- function(x, margin=1) {
  if (!inherits(x, "SpatialPointsDataFrame") & !inherits(x, "SpatialPolygonsDataFrame")) 
    stop("MUST BE sp SpatialPointsDataFrame OR SpatialPolygonsDataFrame CLASS OBJECT") 
  na.index <- unique(as.data.frame(which(is.na(x@data),arr.ind=TRUE))[,margin])
  if(margin == 1) {  
    cat("DELETING ROWS: ", na.index, "\n") 
    return( x[-na.index,]  ) 
  }
  if(margin == 2) {  
    cat("DELETING COLUMNS: ", na.index, "\n") 
    return( x[,-na.index]  ) 
  }
}

## extract_krige_stats
# Extracts summary statistics from kriging cross-validation
extract_krige_stats <- function(krige_model) {
  stat_table <- data.frame(me = NA,
                           mspe = NA,
                           msne = NA,
                           cor_obspred = NA,
                           cor_predres = NA,
                           rmse = NA)
  krige_model <- drop_na(as.data.frame(krige_model))
  stat_table[1, "me"] <- mean(krige_model$residual)
  stat_table[1, "mspe"] <- mean(krige_model$residual^2)
  stat_table[1, "msne"] <- mean(krige_model$zscore^2)
  stat_table[1, "cor_obspred"] <- cor(krige_model$observed, krige_model$var1.pred)
  stat_table[1, "cor_predres"] <- cor(krige_model$var1.pred, krige_model$residual)
  stat_table[1, "rmse"] <- sqrt(mean((krige_model$observed - krige_model$var1.pred)^2))
  return(stat_table)
}

```

## Set paths and read data

Next we set the paths to the project folder, and the various files within it:

```{r set paths}
# Set path to project
projdir <- paste0(here(), "/")

# Set paths to geodatabases
birkhowe.gdb <- paste0(projdir, "MikeBirkhowe.gdb")
hollins.gdb <- paste0(projdir, "MikeHollins.gdb")
lowsnab.gdb <- paste0(projdir, "MikeLowsnab.gdb")

# Examine Feature Classes present in geodatabases
(birkhowe_fclist <- ogrListLayers(birkhowe.gdb))
(hollins_fclist <- ogrListLayers(hollins.gdb))
(lowsnab_fclist <- ogrListLayers(lowsnab.gdb))
```

We need to read the Feature Classes containing the point data for the analyses from the geodatabases:

```{r read features}
# Read Feature Classes from geodatabases
# One per depth, three depths per site
birkhowe_d1 <- readOGR(dsn = birkhowe.gdb, layer = "birkhowe_soild1_pt_v5", pointDropZ = TRUE)
birkhowe_d2 <- readOGR(dsn = birkhowe.gdb, layer = "birkhowe_soild2_pt_v5", pointDropZ = TRUE)
birkhowe_d3 <- readOGR(dsn = birkhowe.gdb, layer = "birkhowe_soild3_pt_v5", pointDropZ = TRUE)
hollins_d1 <- readOGR(dsn = hollins.gdb, layer = "hollins_soild1_pt_v3", pointDropZ = TRUE)
hollins_d2 <- readOGR(dsn = hollins.gdb, layer = "hollins_soild2_pt_v3", pointDropZ = TRUE)
hollins_d3 <- readOGR(dsn = hollins.gdb, layer = "hollins_soild3_pt_v3", pointDropZ = TRUE)
lowsnab_d1 <- readOGR(dsn = lowsnab.gdb, layer = "lowsnab_soild1_pt_v5", pointDropZ = TRUE)
lowsnab_d2 <- readOGR(dsn = lowsnab.gdb, layer = "lowsnab_soild2_pt_v5", pointDropZ = TRUE)
lowsnab_d3 <- readOGR(dsn = lowsnab.gdb, layer = "lowsnab_soild3_pt_v5", pointDropZ = TRUE)
```

We also need to read the elevation data for Hollins, which was missing from the Feature Class in the Hollins geodatabase:

```{r read Hollins elevation}
# Read elevation data for Hollins (missing from geodatabase)
hollins_elevation <- read_csv(paste0(projdir, "hollins_elevation.csv"))
```

## Join data sources ready for analyses

The elevation data for Hollins needs to be joined to the existing data for each of the depths sampled at Hollins. We can use this opportunity to do some additional tidying of the data for consistency with the Lowsnab and Birkhowe data.

```{r join Hollins elevation}
# Remove empty elevation column from spatial data for Hollins
# Remove column hls_plot and rename hls_vegcomm to hls_plot, for consistency with other
# data
# Left-join the elevation data to the existing data for each depth
hollins_d1@data <- select(hollins_d1@data, -elevation, -hls_plot) %>%
                    rename(hls_plot = hls_vegcomm) %>%
                    left_join(hollins_elevation, by = "gps_id")
hollins_d2@data <- select(hollins_d2@data, -elevation, -hls_plot) %>%
                    rename(hls_plot = hls_vegcomm) %>%
                    left_join(hollins_elevation, by = "gps_id")
hollins_d3@data <- select(hollins_d3@data, -elevation, -hls_plot) %>%
                    rename(hls_plot = hls_vegcomm) %>%
                    left_join(hollins_elevation, by = "gps_id")
```

We prepare the data for analysis by removing columns that aren't specified in the model formula and removing NAs, using the function defined above. We do this for every site and depth apart from Lowsnab, depth 1, because Lowsnab doesn't contain an NAs and this causes the function to return an empty dataframe.

```{r select columns and remove NA}
# Subset necessary columns from data
sub_cols <- c("hls_plot", "elevation", "moisture", "totC_mean_mass_vol")
birkhowe_d1_sub <- birkhowe_d1[, sub_cols]
birkhowe_d2_sub <- birkhowe_d2[, sub_cols]
birkhowe_d3_sub <- birkhowe_d3[, sub_cols]

hollins_d1_sub <- hollins_d1[, sub_cols]
hollins_d2_sub <- hollins_d2[, sub_cols]
hollins_d3_sub <- hollins_d3[, sub_cols]

lowsnab_d1_sub <- lowsnab_d1[, sub_cols]
lowsnab_d2_sub <- lowsnab_d2[, sub_cols]
lowsnab_d3_sub <- lowsnab_d3[, sub_cols]

# Filter rows containing NAs and remove duplicates
# Overwrite _sub object
birkhowe_d1_sub <- sp.na.omit(remove.duplicates(birkhowe_d1_sub))
birkhowe_d2_sub <- sp.na.omit(remove.duplicates(birkhowe_d2_sub))
birkhowe_d3_sub <- sp.na.omit(remove.duplicates(birkhowe_d3_sub))
hollins_d1_sub <- sp.na.omit(remove.duplicates(hollins_d1_sub))
hollins_d2_sub <- sp.na.omit(remove.duplicates(hollins_d2_sub))
hollins_d3_sub <- sp.na.omit(remove.duplicates(hollins_d3_sub))
# lowsnab_d1_sub <- sp.na.omit(lowsnab_d1_sub) # No empty cells. Function returns empty data
lowsnab_d2_sub <- sp.na.omit(remove.duplicates(lowsnab_d2_sub))
lowsnab_d3_sub <- sp.na.omit(remove.duplicates(lowsnab_d3_sub))
```

## Fit variograms using gstat::fit.variogram

There are a couple of methods for fitting variograms to data. The fit.variogram procedure in gstat will attempt to optimise and select an appropriate variogram model. We use gstat::fit.variogram to fit a variogram for each site and depth to start with:

```{r gstat::fit.variogram, echo=FALSE}
birkhowe_d1.vgmfit <- fit.variogram(variogram(totC_mean_mass_vol ~ 1, birkhowe_d1_sub),
                                    vgm(c("Exp", "Mat", "Sph", "Ste", "Gau")),
                                    fit.kappa = TRUE,
                                    debug.level = 1)

birkhowe_d2.vgmfit <- fit.variogram(variogram(totC_mean_mass_vol ~ 1, birkhowe_d2_sub),
                                    vgm(c("Exp", "Mat", "Sph", "Ste", "Gau")),
                                    fit.kappa = TRUE,
                                    debug.level = 1)

birkhowe_d3.vgmfit <- fit.variogram(variogram(totC_mean_mass_vol ~ 1, birkhowe_d3_sub),
                                    vgm(c("Exp", "Mat", "Sph", "Ste", "Gau")),
                                    fit.kappa = TRUE,
                                    debug.level = 1)

lowsnab_d1.vgmfit <- fit.variogram(variogram(totC_mean_mass_vol ~ 1, lowsnab_d1_sub),
                                    vgm(c("Exp", "Mat", "Sph", "Ste", "Gau")),
                                    fit.kappa = TRUE,
                                   debug.level = 1)

lowsnab_d2.vgmfit <- fit.variogram(variogram(totC_mean_mass_vol ~ 1, lowsnab_d2_sub),
                                   vgm(c("Exp", "Mat", "Sph", "Ste", "Gau")),
                                   fit.kappa = TRUE,
                                   debug.level = 1)

lowsnab_d3.vgmfit <- fit.variogram(variogram(totC_mean_mass_vol ~ 1, lowsnab_d3_sub),
                                   vgm(c("Exp", "Mat", "Sph", "Ste", "Gau")),
                                   fit.kappa = TRUE,
                                   debug.level = 1)

hollins_d1.vgmfit <- fit.variogram(variogram(totC_mean_mass_vol ~ 1, hollins_d1_sub),
                                   vgm(c("Exp", "Mat", "Sph", "Ste", "Gau")),
                                   fit.kappa = TRUE,
                                   debug.level = 1)

hollins_d2.vgmfit <- fit.variogram(variogram(totC_mean_mass_vol ~ 1, hollins_d2_sub),
                                   vgm(c("Exp", "Mat", "Sph", "Ste", "Gau")),
                                   fit.kappa = TRUE,
                                   debug.level = 1)

hollins_d3.vgmfit <- fit.variogram(variogram(totC_mean_mass_vol ~ 1, hollins_d3_sub),
                                   vgm(c("Exp", "Mat", "Sph", "Ste", "Gau")),
                                   fit.kappa = TRUE,
                                   debug.level = 1)
```

## Fit variograms using automap::autofitVariogram

Another option available to us for fitting variograms is the automap::autofitVariogram function. This is a wrapper for gstat::fit.variogram, but uses a different set of starting values to fit variograms and a different selection procedure to select the optimal variogram. This might yield different results to gstat::fit.variogram, so we repeat the steps above using this procedure. The function reports the optimal variogram selected and details of the optimisation procedure.

```{r automap::autofitVariogram, echo=FALSE}
birkhowe_d1.vgmafit <- autofitVariogram(totC_mean_mass_vol ~ 1,
                                        birkhowe_d1_sub,
                                        verbose = TRUE)

birkhowe_d2.vgmafit <- autofitVariogram(totC_mean_mass_vol ~ 1,
                                        birkhowe_d2_sub,
                                        verbose = TRUE)

birkhowe_d3.vgmafit <- autofitVariogram(totC_mean_mass_vol ~ 1,
                                        birkhowe_d3_sub,
                                        verbose = TRUE)

lowsnab_d1.vgmafit <- autofitVariogram(totC_mean_mass_vol ~1,
                                       lowsnab_d1_sub,
                                       verbose = TRUE)

lowsnab_d2.vgmafit <- autofitVariogram(totC_mean_mass_vol ~ 1,
                                       lowsnab_d2_sub,
                                       verbose = TRUE)

lowsnab_d3.vgmafit <- autofitVariogram(totC_mean_mass_vol ~ 1,
                                       lowsnab_d3_sub,
                                       verbose = TRUE)

hollins_d1.vgmafit <- autofitVariogram(totC_mean_mass_vol ~ 1,
                                       hollins_d1_sub,
                                       verbose = TRUE)

hollins_d2.vgmafit <- autofitVariogram(totC_mean_mass_vol ~ 1,
                                       hollins_d2_sub,
                                       verbose = TRUE)

hollins_d3.vgmafit <- autofitVariogram(totC_mean_mass_vol ~ 1,
                                       hollins_d3_sub,
                                       verbose = TRUE)
```

## Kriging with cross-validation

With the variograms set up using two different methods, we can proceed with the regression kriging. We krige using each variogram for each site and depth, to see which variogram produces the best fit.

N-fold cross validation is used. This is a leave-one-out procedure - the modelled surface is fitted for as many data points that there are at each site, -1, leaving one data point out each time.

```{r kriging with n-fold cross validation, echo=FALSE}
# Birkhowe D1
birkhowe_d1.cv <- krige.cv(formula = totC_mean_mass_vol ~ elevation + moisture + hls_plot,
                           birkhowe_d1_sub,
                           model = birkhowe_d1.vgmfit,
                           debug.level = 2)

(birkhowe_d1.stats <- extract_krige_stats(birkhowe_d1.cv))

birkhowe_d1.afcv <- krige.cv(formula = totC_mean_mass_vol ~ elevation + moisture + hls_plot,
                           birkhowe_d1_sub,
                           model = birkhowe_d1.vgmafit$var_model,
                           debug.level = 2)

(birkhowe_d1.astats <- extract_krige_stats(birkhowe_d1.afcv))

# Birkhowe d2
birkhowe_d2.cv <- krige.cv(formula = totC_mean_mass_vol ~ elevation + moisture + hls_plot,
                           birkhowe_d2_sub,
                           model = birkhowe_d2.vgmfit,
                           debug.level = 2)

(birkhowe_d2.stats <- extract_krige_stats(birkhowe_d2.cv))

birkhowe_d2.afcv <- krige.cv(formula = totC_mean_mass_vol ~ elevation + moisture + hls_plot,
                           birkhowe_d2_sub,
                           model = birkhowe_d2.vgmafit$var_model,
                           debug.level = 2)

(birkhowe_d2.astats <- extract_krige_stats(birkhowe_d2.afcv))

# Birkhowe d3
birkhowe_d3.cv <- krige.cv(formula = totC_mean_mass_vol ~ elevation + moisture + hls_plot,
                           birkhowe_d3_sub,
                           model = birkhowe_d3.vgmfit,
                           debug.level = 2)

(birkhowe_d3.stats <- extract_krige_stats(birkhowe_d3.cv))

birkhowe_d3.afcv <- krige.cv(formula = totC_mean_mass_vol ~ elevation + moisture + hls_plot,
                           birkhowe_d3_sub,
                           model = birkhowe_d3.vgmafit$var_model,
                           debug.level = 2)

(birkhowe_d3.astats <- extract_krige_stats(birkhowe_d3.afcv))

# Lowsnab d1
lowsnab_d1.cv <- krige.cv(formula = totC_mean_mass_vol ~ elevation + moisture + hls_plot,
                          lowsnab_d1_sub,
                          model = lowsnab_d1.vgmfit,
                          debug.level = 2)

(lowsnab_d1.stats <- extract_krige_stats(lowsnab_d1.cv))

lowsnab_d1.afcv <- krige.cv(formula = totC_mean_mass_vol ~ elevation + moisture + hls_plot,
                          lowsnab_d1_sub,
                          model = lowsnab_d1.vgmafit$var_model,
                          debug.level = 2)

(lowsnab_d1.astats <- extract_krige_stats(lowsnab_d1.afcv))

# Lowsnab D2
lowsnab_d2.cv <- krige.cv(formula = totC_mean_mass_vol ~ elevation + moisture + hls_plot,
                          lowsnab_d2_sub,
                          model = lowsnab_d2.vgmfit,
                          debug.level = 2)

(lowsnab_d2.stats <- extract_krige_stats(lowsnab_d2.cv))

lowsnab_d2.afcv <- krige.cv(formula = totC_mean_mass_vol ~ elevation + moisture + hls_plot,
                          lowsnab_d2_sub,
                          model = lowsnab_d2.vgmafit$var_model,
                          debug.level = 2)

(lowsnab_d2.astats <- extract_krige_stats(lowsnab_d2.afcv))

# Lowsnab D3
lowsnab_d3.cv <- krige.cv(formula = totC_mean_mass_vol ~ elevation + moisture + hls_plot,
                          lowsnab_d3_sub,
                          model = lowsnab_d3.vgmfit,
                          debug.level = 2)

(lowsnab_d3.stats <- extract_krige_stats(lowsnab_d3.cv))

lowsnab_d3.afcv <- krige.cv(formula = totC_mean_mass_vol ~ elevation + moisture + hls_plot,
                          lowsnab_d3_sub,
                          model = lowsnab_d3.vgmafit$var_model,
                          debug.level = 2)

(lowsnab_d3.astats <- extract_krige_stats(lowsnab_d3.afcv))

# Hollins D1
hollins_d1.cv <- krige.cv(formula = totC_mean_mass_vol ~ elevation + moisture + hls_plot,
                          hollins_d1_sub,
                          model = hollins_d1.vgmfit,
                          debug.level = 2)

(hollins_d1.stats <- extract_krige_stats(hollins_d1.cv))

hollins_d1.afcv <- krige.cv(formula = totC_mean_mass_vol ~ elevation + moisture + hls_plot,
                          hollins_d1_sub,
                          model = hollins_d1.vgmafit$var_model,
                          debug.level = 2)

(hollins_d1.astats <- extract_krige_stats(hollins_d1.afcv))

# Hollins D2
hollins_d2.cv <- krige.cv(formula = totC_mean_mass_vol ~ elevation + moisture + hls_plot,
                          hollins_d2_sub,
                          model = hollins_d2.vgmfit,
                          debug.level = 2)

(hollins_d2.stats <- extract_krige_stats(hollins_d2.cv))

hollins_d2.afcv <- krige.cv(formula = totC_mean_mass_vol ~ elevation + moisture + hls_plot,
                          hollins_d2_sub,
                          model = hollins_d2.vgmafit$var_model,
                          debug.level = 2)

(hollins_d2.astats <- extract_krige_stats(hollins_d2.afcv))

# Hollins D3
hollins_d3.cv <- krige.cv(formula = totC_mean_mass_vol ~ elevation + moisture + hls_plot,
                          hollins_d3_sub,
                          model = hollins_d3.vgmfit,
                          debug.level = 2)

(hollins_d3.stats <- extract_krige_stats(hollins_d3.cv))

hollins_d3.afcv <- krige.cv(formula = totC_mean_mass_vol ~ elevation + moisture + hls_plot,
                          hollins_d3_sub,
                          model = hollins_d3.vgmafit$var_model,
                          debug.level = 2)

(hollins_d3.astats <- extract_krige_stats(hollins_d3.afcv))
```

## Output model statistics

Consolidate kriging summaries into site tables

```{r kriging summary tables}
birkhowe.stats <- bind_rows(list(birkhowe_d1.stats,
                                 birkhowe_d2.stats,
                                 birkhowe_d3.stats),
                            .id = "depth")

lowsnab.stats <- bind_rows(list(lowsnab_d1.stats,
                                lowsnab_d2.stats,
                                lowsnab_d3.stats),
                           .id = "depth")

hollins.stats <- bind_rows(list(hollins_d1.stats,
                                hollins_d2.stats,
                                hollins_d3.stats),
                           .id = "depth")

birkhowe.astats <- bind_rows(list(birkhowe_d1.astats,
                                 birkhowe_d2.astats,
                                 birkhowe_d3.astats),
                            .id = "depth")

lowsnab.astats <- bind_rows(list(lowsnab_d1.astats,
                                lowsnab_d2.astats,
                                lowsnab_d3.astats),
                           .id = "depth")

hollins.astats <- bind_rows(list(hollins_d1.astats,
                                hollins_d2.astats,
                                hollins_d3.astats),
                           .id = "depth")
```

We combine the summary stats for each site and depth into two tables - one for each kriging method.

**gstat::fit.variogram method:**

```{r overall summary tables - gstat}
allsite.stats <- bind_rows(list(birkhowe = birkhowe.stats,
                                lowsnab = lowsnab.stats,
                                hollins = hollins.stats),
                           .id = "site")
```
                  
```{r gstat krige fits summary table, echo=FALSE, results="asis"}
kable(allsite.stats, caption = "Kriging model fits, gstat::fit.variogram method")
```

**automap::autofitVariogram method:**

```{r overall summary tables - automap}
allsite.astats <- bind_rows(list(birkhowe = birkhowe.astats,
                                 lowsnab = lowsnab.astats,
                                 hollins = hollins.astats),
                            .id = "site")
```

```{r automap krige fits summary table, echo=FALSE, results="asis"}
kable(allsite.astats, caption = "Kriging model fits, automap::autofitVariogram method")
```
